Adversarial deep neural networks remove nonlinear batch effects from gene-expression data

### 2019-07-16 manuscript.tex
Renamed file

Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
However, these methods are designed for a limited range of scenarios:
1) when the input data contain only two batches,
2) the batches are sufficiently large (batch sizes smaller than $n=100$ failed in our testing), and
3) the batches are balanced.[@For \#2, does this mean the total number of samples? Or the number of samples per batch? Please clarify the wording. -Is this more clear?@]
These requirements rarely hold in applied research; for example, the bladderbatch dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
Additionally, existing methods for batch correction have not been tested to verify whether nonlinear patterns remain in expression data after batch correction.
To address these limitations, we implemented a deep neural network that 1) has a batch output layer that expands to fit
the number of batches, 2) works on smaller batch sizes by sampling with replacement from the training data, and 3) has
no formulas that require batches to be equal sizes. [@describe our method briefly, what is different about our method, and why we think our method can be applied when there are >2 batches, small datasets, and unbalanced batches. Is this clear enough? @]
Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remained after adjustment.

Autoencoders are a type of neural network that encodes input data in fewer dimensions and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}[@Please verify that this is a correct reference to use here. - This reference looks great to me.@].
This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit. - Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise? Either way, we might need a reference.@]
Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
However, neural networks are adept at finding subtle patterns in data; thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.[@Make sure this wording and the references are appropriate. - The wording and both references look great to me.@]
Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions: 1) to maximize similarity between in the input and output data and 2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
Louizos, et al. have used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.

>>>>

Recent progress has been made in using artificial neural networks to correct batch effects in expression data \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019}.
However, these methods are designed for a limited range of scenarios:
1) when the input data contain only two batches,
2) the batches are sufficiently large (batch sizes smaller than 100 failed in our testing), and
3) the batches are balanced.
These requirements rarely hold in applied research; for example, the \textit{bladderbatch} dataset---which is used to test batch-adjustment techniques\cite{leek_bladderbatch_2017,leek_sva_2017}---has 5 batches, 57 samples, and 4 to 19 samples per batch.
Additionally, papers that have described these methods have not evaluated whether nonlinear patterns remain in expression data after batch correction.
To address these limitations, we implemented a deep neural-network architecture that 1) has a batch output layer that expands to fit
the number of batches, 2) works on relatively small batch sizes by sampling with replacement from the training data, and 3) has
no formulas that require batches to be equal sizes.
Furthermore, we tested our method using machine-learning classification algorithms to quantify the extent to which nonlinear batch effects remain after adjustment.

Autoencoders are a type of neural network that encodes input data in a smaller number of dimensions than the original data and then reconstructs the data in the original dimensions\citep{hinton_reducing_2006}.
This technique can be used to reduce noise and thus refine the data representation.[@Make sure this wording is legit. - Are you referencing denoising autoencoders, or are you saying that vanilla autoencoders reduce noise? Either way, we might need a reference. You would know better than I. Please clarify this based on what is relevant to the paper and add a relevant reference.@]
Typically, autoencoders are trained using an objective function that seeks to ensure that the output is as similar as possible to the input \citep{hinton_reducing_2006}.
However, neural networks are adept at finding subtle patterns in data;
thus the objective function may prioritize replication of confounding patterns, such as batch effects, rather than the signals of interest \citep{ganin_domain-adversarial_2015,louizos_causal_2017-2}.
Recently, researchers have experimented with discouraging neural networks from learning such patterns by giving them two competing objective functions:
1) to maximize similarity between in the input and output data and
2) to ignore any patterns associated with known confounding variables \citep{ganin_domain-adversarial_2015,tzeng_deep_2014-2}.
For example, Louizos, et al. used this type of dual-objective function in combination with a variational autoencoder and successfully removed domain-based variability in credit scores, financial savings, and hospital admittance datasets \citet{louizos_variational_2015}.
